A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-04-11 jacobian_calc_CMAME.tex
Updates to MDB changes

\note{Re: I.4} \add{As mentioned earlier, the CS method was originally introduced in a paper by Lyness and Moler} \cite{lyness1967numerical}\add{.  A follow-on contribution by Lyness demonstrated a ``truncated'' approximate version of the original CS method to make it suitable for use on digital computers} \cite{lyness1968differentiation}\add{.  After these original papers, little can be found in the literature regarding the method until Squire and Trapp showed through empirical data that CS offers superior accuracy to FD and CD} \cite{squire1998using} \add{in computing numerical derivatives. Papers by Newman et al.\ } \cite{newman1998} \add{and Anderson et al.\ } \cite{anderson2001sensitivity} \add{discuss computing sensitivities in a CFD analysis using CS and compares the results to sensitivities produced with central-difference derivatives. The work is also notable for identifying automatic-differentiation as a competing method to CS and suggesting that complex-step could be used to compute Jacobians of residuals for use in Newton-Krylov schemes for solving non-linear systems. Perez-Foguet et al.\ } \cite{perez2000numerical, perez2012numerical} \add{used the CS method to compute algorithmically consistent elastoplastic moduli in non-linear solid mechanics problems.  A paper by Joaquim et al.\ } \cite{martins2000automated} \add{investigates using CS for computing aerodynamic sensitivities. Additionally, higher-order accuracy versions of CS are discussed, as is the task of implementation in most widely used general purpose programming languages. Discussed is the relative ease of implementing a CS scheme versus an automatic-differentiation based scheme for computing sensitivities in an established code.  
Martins et al.\ } \cite{martins2003complex}  \add{demonstrates that the CS method can be performed by using complex datatypes for the variables of functions comprising an analysis code. The accuracy of this approach, with respect to the derivatives themselves, is compared to automatic-differentiation in a serial analysis code. Lai and Crassidis} \cite{lai2008extensions} \add{discuss first- and second-order CS methods for computing derivatives and choosing optimal step sizes.  Al-Mohy and Higham} \cite{al2010complex} \add{investigate the use of CS in computing Fr\'{e}chet derivatives of matrix functions. Jin et al.\ } \cite{jin2010improved} \add{discuss using CS to compute sensitivities in a solid mechanics FEM analysis.  Voorhees et al.\ } \cite{voorhees2011complex} \add{discusses the background of the CS method and frames it as a subset of Fourier Differentiation. CS and Fourier Differentiation are then applied to computing sensitivities within a solid mechanics finite element analysis. They implemented the complex-values as additional degrees of freedom in the finite element stiffness matrices using a real-valued matrix representation of the complex numbers where required. In a follow-on work,  Millwater et al.\ } \cite{millwater2013application} \add{extended this concept, calling it ZFEM, and implemented the method into user defined element types for use with the popular ABAQUS} \cite{systemes2012abaqus} \add{FEA software package.  Finally, Abreu et al.\ } \cite{abreu2013generalization} \add{discusses computing derivatives with a version of the CS method which features the combined use of real and imaginary valued step components, as in Fourier differentiation.  They suggests that the modified CS method can achieve extended approximation accuracy up to fourth order.  The body of prior literature has introduced the application of CS to single dimension and multidimensional derivatives, sensitivity analysis, non-linear solution schemes, the extension of CS for higher accuracy, and implementation strategies for CS in new and existing codes.  With respect to the prior work discussing the CS method, the novel aspects of the instant work are then its discussion of the application of CS to peridynamics simulations run in \emph{Peridigm}, a parallel implementation of CS and a compilation of CPU time data, where each of CS, AD, FD and CD are compared.}

@@

The \emph{forward difference} (FD) formula for calculating a tangent-stiffness matrix in the setting of a solid structural mechanics problem is 
%
\begin{equation} 
  K_{ij} \approx \frac{F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u})}{h},
\end{equation}
%

where $K_{ij}$ is the indicial notation representation of an element of the
tangent-stiffness matrix at row $i$, column $j$. The first of the two terms in
the numerator is the internal force, $F_i^{int}$ evaluated in the current
deformed configuration, as a function of the displacement, $\vec{u}$,  plus a small perturbation $h$ in the
direction $\hat{e}_j$ where $\hat{e}_j$ represents a unit-vector corresponding
to the $j^{\mbox{th}}$ degree-of-freedom. The second of the two terms is
$F_i^{int}$ evaluated in the current configuration. The denominator is the
magnitude of perturbation called the \emph{probe-distance}.

@@

\note[]{Re: I.5, next section is taken from the thesis.}
\section{Working examples of CS, AD and FD for Newton's method}
\label{sec:WorkingExamples}

In order to illustrate the complex-step technique in a simpler form than it
appears in the \emph{Peridigm} code written for this study, an example
header-only library and two example problem definition programs that use it
have been created. All the materials described here are available in the
associated project repository.

@@

The source code for the header-only library is entitled
\emph{NewtonRaphson.hpp}. Note that the library is an interface to functions and
data structures made available in the \emph{Trilinos} library, mentioned
earlier, such that it is required in order for the examples described here
to work. \emph{Trilinos} is freely available for download as of this
writing at: \emph{trilinos.sandia.gov}.  Generally the header-only library
works by taking in the users initial guessed equilibrium solution, target
dependent variable values, system constants, and a function pointer to a
model evaluation function. Solver parameters including an update scaling
factor, tolerance, maximum iterations and probe length are chosen after the
solver is instantiated as a new object. The behavior of the solver is
specialized through a coding feature called inheritance to perform each of
CS, AD and FD.  But, in common, each version of the solver works by
computing the system Jacobian, solving a linear system as in that step of
the Newton Raphson method, computing the residual and checking for
convergence or overrun. Here the linear solution method is LU
decomposition, while the chief convergence criterion is change in residual
between iterations.  Additionally the problem can be partially respecified
to allow for running a series of load steps having progressing target
values, which is a typical mode also seen in \emph{Peridigm}. Applying
boundary conditions incrementally in load steps assists convergence
\cite{rezaiee2010dynamic}.

\subsection{First example}
\label{subsec:ProgOne}
The first example program, \emph{NRexamples.cpp}, solves a simple nonlinear
system describing the intersection of three infinite paraboloids in threespace.
It works by defining force computation functions for each of the models, and
then passing an address called a function pointer corresponding to each of
these methods to a solver object which coordinates solving the problem and
reporting output for the user based on their convergence criteria and other
preferences like step-size and maximum number of iterations. CS, AD and FD are
each represented and the user is free to adjust input settings by specifying
command line arguments as the compiled program instructs, or perhaps create a
shell script that varies parameters for each of the methods in order to perform
a comparative study.

\subsection{Second example and related studies} 
\label{subsec:Validation}
The second program,a \emph{NRComparison.cpp}, follows a case study that appears in
\cite{rezaiee2010dynamic}. The example program departs from the cited
literature by applying each of CS, AD and FD and verifying them against an
analytical model of a nonlinear mechanical system involving a connected beam
and spring. Fig. \ref{fig:TrussSchematic} is a schematic of the beam and spring
system.

\begin{figure}[tbp] \centering \includegraphics[width=0.7\textwidth]{./figs/truss.png}
\caption{Beam and spring simulated in example program.} \label{fig:TrussSchematic}
\end{figure}

The equations describing this system and results for verification were obtained
in \cite{rezaiee2010dynamic}. They include a force function, Eqn.
\ref{eqn:TrussForce}, and an analytical force derivative or stiffness, Eqn.
\ref{eqn:TrussStiffness}.

>>>>

\note{Re: I.4} \add{As mentioned earlier, the CS method was originally introduced in a paper by Lyness and Moler} \cite{lyness1967numerical}\add{.  A follow-on contribution by Lyness demonstrated a ``truncated'' approximate version of the original CS method to make it suitable for use on digital computers} \cite{lyness1968differentiation}\add{.  After these original papers, little can be found in the literature regarding the method until Squire and Trapp showed through numerical experiments that CS offers superior accuracy to FD and CD} \cite{squire1998using} \add{in computing numerical derivatives. Papers by Newman et al.\ } \cite{newman1998} \add{and Anderson et al.\ } \cite{anderson2001sensitivity} \add{discuss computing sensitivities in a CFD analysis using CS and compares the results to sensitivities produced with central-difference derivatives. The work is also notable for identifying automatic-differentiation as a competing method to CS and suggesting that complex-step could be used to compute Jacobians of residuals for use in Newton-Krylov schemes for solving non-linear systems. Perez-Foguet et al.\ } \cite{perez2000numerical, perez2012numerical} \add{used the CS method to compute algorithmically consistent elastoplastic moduli in non-linear solid mechanics problems.  A paper by Joaquim et al.\ } \cite{martins2000automated} \add{investigates using CS for computing aerodynamic sensitivities. Additionally, higher-order accuracy versions of CS are discussed, as is the task of implementation in most widely used general purpose programming languages. Discussed is the relative ease of implementing a CS scheme versus an automatic-differentiation based scheme for computing sensitivities in an established code.  
Martins et al.\ } \cite{martins2003complex}  \add{demonstrates that the CS method can be performed by using complex datatypes for the variables of functions comprising an analysis code. The accuracy of this approach, with respect to the derivatives themselves, is compared to automatic-differentiation in a serial analysis code. Lai and Crassidis} \cite{lai2008extensions} \add{discuss first- and second-order CS methods for computing derivatives and choosing optimal step sizes.  Al-Mohy and Higham} \cite{al2010complex} \add{investigate the use of CS in computing Fr\'{e}chet derivatives of matrix functions. Jin et al.\ } \cite{jin2010improved} \add{discuss using CS to compute sensitivities in a solid mechanics FEM analysis.  Voorhees et al.\ } \cite{voorhees2011complex} \add{discusses the background of the CS method and frames it as a subset of Fourier Differentiation. CS and Fourier Differentiation are then applied to computing sensitivities within a solid mechanics finite element analysis. They implemented the complex-values as additional degrees of freedom in the finite element stiffness matrices using a real-valued matrix representation of the complex-numbers where required. In a follow-on work,  Millwater et al.\ } \cite{millwater2013application} \add{extended this concept, calling it ZFEM, and implemented the method into user defined element types for use with the popular ABAQUS} \cite{systemes2012abaqus} \add{FEA software package.  Nishida et al.\ } \cite{nishida2013} \add{used the complex-step method to compute Jacobian matrices for implementation in a Levenberg-Marquardt optimization routine for design-of-experiments purposes.  Finally, Abreu et al.\ } \cite{abreu2013generalization} \add{discusses computing derivatives with a version of the CS method which features the combined use of real and imaginary valued step components, as in Fourier differentiation.  They suggests that the modified CS method can achieve extended approximation accuracy up to fourth order.  This recent literature has introduced the application of CS to single-dimension and multidimensional derivatives, sensitivity analysis, non-linear solution schemes, optimization, the extension of CS for higher accuracy, and implementation strategies for CS in new and existing codes.  The current paper presents a comparison of the complex-step method to the other methods for numerical differentiation in as close to a one-to-one implementation as possible for direct comparison purposes in a large-scale engineering analysis code. To the author's knowledge it is the first application of the complex-step method to solve peridynamics mechanics problems, and the first implementation and comparison of the method in large-scale distributed memory parallel setting. }

@@

The \emph{forward difference} (FD) formula for calculating a tangent-stiffness matrix in the setting of a solid structural mechanics problem is 
%
\begin{equation} 
  K_{ij} \approx \frac{F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u})}{h},
\end{equation}
%
where $K_{ij}$ is the indicial notation representation of an element of the tangent-stiffness matrix at row $i$, column $j$. The first of the two terms in the numerator is the internal force, $F_i^{int}$ evaluated in the current deformed configuration, as a function of the displacement, $\vec{u}$,  plus a small perturbation $h$ in the direction $\hat{e}_j$ where $\hat{e}_j$ represents a unit-vector corresponding to the $j^{\mbox{th}}$ degree-of-freedom. The second of the two terms is $F_i^{int}$ evaluated in the current configuration. The denominator is the
magnitude of perturbation called the \emph{probe-distance}.

@@

\note[]{Re: I.5, next section is taken from the thesis.}
\section{Simple example of CS, AD and FD for solution of nonlinear problems.}
\label{sec:WorkingExamples}

In order to illustrate a comparison of the numerical differentiation techniques in a simpler form than it appears in the large-scale analysis code covered in the following sections, a simple example borrowed from \cite{rezaiee2010dynamic} was implemented with each of the methods, CS, AD, and FD. All the source code to reproduce the results described here are available in the associated paper repository.

@@

The source code for the header-only library is entitled \emph{NewtonRaphson.hpp}. Note that the library is an interface to functions and data structures made available in the \emph{Trilinos} library, mentioned earlier, such that it is required in order for the examples described here to work. \emph{Trilinos} is freely available for download as of this writing at: \emph{trilinos.sandia.gov}.  Generally the header-only library works by taking in the users initial guessed equilibrium solution, target dependent variable values, system constants, and a function pointer to a model evaluation function. Solver parameters including an update scaling factor, tolerance, maximum iterations and probe length are chosen after the solver is instantiated as a new object. The behavior of the solver is specialized through a coding feature called inheritance to perform each of CS, AD and FD.  But, in common, each version of the solver works by computing the system Jacobian, solving a linear system as in that step of the Newton Raphson method, computing the residual and checking for convergence or overrun. Here the linear solution method is LU decomposition, while the chief convergence criterion is change in residual between iterations.  Additionally the problem can be partially respecified to allow for running a series of load steps having progressing target values, which is a typical mode also seen in \emph{Peridigm}. Applying boundary conditions incrementally in load steps assists convergence.

\subsection{Second example and related studies} 
\label{subsec:Validation}
The second program,a \emph{NRComparison.cpp}, follows a case study that appears in \cite{rezaiee2010dynamic}. The example program departs from the cited literature by applying each of CS, AD and FD and verifying them against an analytical model of a nonlinear mechanical system involving a connected beam and spring. Fig. \ref{fig:TrussSchematic} is a schematic of the beam and spring system. 
%
\begin{figure}[tbp] \centering \includegraphics[width=0.7\textwidth]{./figs/truss.png}
\caption{Beam and spring simulated in example program.} \label{fig:TrussSchematic}
\end{figure}
%
The equations describing this system and results for verification were obtained
in \cite{rezaiee2010dynamic}. They include a force function, Eqn.
\ref{eqn:TrussForce}, and an analytical force derivative or stiffness, Eqn.
\ref{eqn:TrussStiffness}.
